Perturbing SU($n$)-symmetric interactions

### 2020-01-16 collective_sun/sun_perturbation.tex
changed notation for denoting collective opetrators

\newcommand{\FS}{\text{FS}}
\newcommand{\EQFS}{=_{\text{FS}}}
\newcommand{\col}{\overline}

\DeclareMathOperator{\sign}{sign}
\DeclareMathOperator{\cov}{cov}
\let\var\oldvar
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\diag}{diag}
\let\flat\oldflat
\DeclareMathOperator{\flat}{flat}
\DeclareMathOperator{\counts}{counts}
\newcommand{\flatvec}{\underline\flat\,}
\newcommand{\ul}{\underline}

@@

We now return to the task of computing the second-order effective
Hamiltonian $H_1^{(2)}$.  Any coefficient vector $\v v_X$ can be
expanded into its projections $\v v_X^\Delta$ onto the eigenspace of
vectors satisfying \eqref{eq:cond_1} with eigenvalue $\Delta$,
i.e.~$\v v_X = \sum_\Delta \v v_X^\Delta$, which also allows us to
expand
\begin{align}
  \V_1 = \sum_\Delta \V_1^\Delta,
  &&
  \V_1^\Delta \equiv \sum_X \v v_X^\Delta \c \v X,
\end{align}
where each operator $\V_1^\Delta$ generates a state with definite
interaction energy $\Delta$ above that of the fully symmetric manifold
$\M_0$.  We can therefore simplify
\begin{align}
  H_1^{(2)}
  = - \P_0 \V_1 \E \V_1 \P_0
  = - \sum_{\Delta\ne0} \f1{\Delta}
  \P_0 \V_1^\Delta \P_\Delta \V_1^\Delta \P_0
  = - \sum_{\Delta\ne0} \f1{\Delta} \P_0 \p{\V_1^\Delta}^2 \P_0,
\end{align}
where the product $\P_0 \p{\V_1^\Delta}^2 \P_0$ is worked out in
Appendix \ref{sec:PXYP}.  In total, we find
\begin{align}
  H_1^{(2)}
  \EQFS \f1{N\p{N-1}} \sum_{X,Y} \sum_{\Delta\ne0}
  \f{\v v_X^\Delta\c\v v_Y^\Delta}{\Delta}
  \p{\col{X}\,\col{Y} - N\, \col{XY}},
  \label{eq:H_1_2}
\end{align}
where $\EQFS$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$; $\v v_X^\Delta$ is the projection of
$\v v_X$ onto the $\Delta$-eigenspace of the matrix $\m h-\diag\v h$;
and for any single-body operator $Z$, we define the collective operator
\begin{align}
  \col{Z} \equiv \sum_p Z_p.
\end{align}

@@

Calculating the second-order effective Hamiltonian $H_1^{(2)}$ induced
on the fully symmetric manifold $\M_0$ by the perturbation $\V_1$ in
\eqref{eq:perturbations} requires us to simplify
\begin{align}
  \P_0 \p{\V_1^\Delta}^2 \P_0
  \EQFS \sum_{\substack{X,Y\\p,q}} v_{Xp} v_{Yq} X_p Y_q
  = \sum_{X,Y,p} v_{Xp} v_{Yp} X_p Y_p
  + \sum_{\substack{X,Y\\p\ne q}} v_{Xp} v_{Yq} X_p Y_q,
  \label{eq:PXYP_start}
\end{align}
where $\EQFS$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$; and all coefficient vectors $\v v_X$
satisfy \eqref{eq:cond_1} with eigenvalue $\Delta\ne0$, which implies
that all $\v v_X$ are mean-zero.  The first sum on the right of
\eqref{eq:PXYP_start} is then
\begin{align}
  \sum_{X,Y,p} v_{Xp} v_{Yp} X_p Y_p
  \EQFS \sum_{X,Y,p} v_{Xp} v_{Yp} X_0 Y_0
  \EQFS \f1N \sum_{X,Y} \v v_X \c\v v_Y \col{XY},
  &&
  \col{Z} \equiv \sum_p Z_p.
  \label{eq:PXYP_eq}
\end{align}
The second sum on the right of \eqref{eq:PXYP_start}, meanwhile, is
\begin{align}
  \sum_{\substack{X,Y\\p\ne q}} v_{Xp} v_{Yq} X_p Y_q
  \EQFS \sum_{\substack{X,Y\\p\ne q}} v_{Xp} v_{Yq} X_0 Y_1,
  \label{eq:PXYP_neq_start}
\end{align}
where we can use the fact that the vectors $\v v_X,\v v_Y$ are
mean-zero to simplify the sum over $p\ne q$ in
\eqref{eq:PXYP_neq_start} to
\begin{align}
  \sum_{p\ne q} v_{Xp} v_{Yq}
  = \sum_{p,q} v_{Xp} v_{Yq} - \sum_{p=q} v_{Xp} v_{Yq}
  = - \v v_X\c\v v_Y,
\end{align}
In order to simplify the product of operators in
\eqref{eq:PXYP_neq_start}, we expand
\begin{align}
  \col{X}\,\col{Y} = \sum_{p,q} X_p Y_q
  = \sum_{p=q} X_p Y_q + \sum_{p\ne q} X_p Y_q
  \EQFS \col{XY} + N\p{N-1} X_0 Y_1,
  \label{eq:PXYP_neq_ops}
\end{align}
which implies that
\begin{align}
  \sum_{\substack{X,Y\\p\ne q}} v_{Xp} v_{Yq} X_p Y_q
  \EQFS - \sum_{X\ne Y} \v v_X\c\v v_Y X_0 Y_1
  \EQFS - \f1{N\p{N-1}}\sum_{X,Y} \v v_X\c\v v_Y
  \p{\col{X}\,\col{Y} - \col{XY}}.
  \label{eq:PXYP_neq}
\end{align}
Altogether, we have that
\begin{align}
  \P_0 \p{\V_1^\Delta}^2 \P_0
  = - \f1{N\p{N-1}} \sum_{X,Y} \v v_X \c\v v_Y
  \p{\col{X}\,\col{Y} - N\,\col{XY}}.
\end{align}

@@

  \ul{\check{\m h}}

@@

  \ul{\v{\m w}} \equiv \sum_{d\ne0} w_{d,0} \ket{d},

@@

  \ul{\check{\m h}} \c \ul{\v{\m w}}
  = \p{\f12\Delta + h_0} \ul{\v{\m w}}.

@@

$\ul{\v{\m w}}$ as

@@

  \ul{\v{\m w}} = \sum_{d\in\VV_N} w_{d,0} \,

@@

Finally, we project $\ul{\check{\m h}}$ onto the reflection-symmetric

@@

  \P_+ \ul{\check{\m h}} \P_+

@@

  \P_+ \ul{\check{\m h}} \P_+ \c \ul{\v{\m w}}
  = \p{\f12\Delta + h_0} \ul{\v{\m w}}.

@@

$\P_+ \ul{\check{\m h}} \P_+$ only matter when $N$ is even, in which

@@

$\P_+ \ul{\check{\m h}} \P_+$ by a factor of $1/\sqrt{2}$.

@@

where we place a bar over a single-spin operator to denote its
collective version, i.e.~$\col{Q}\equiv\sum_j Q^{(j)}$, where
$Q^{(j)}$ is an operator that acts with the single-spin operator $Q$
on spin $j$ and trivially on all other spins.  By construction, the
set of collective operators $\set{\O_P:\t{partitions}~P~\t{of}~S}$ is
one-to-one to the set of multi-local operators
$\set{O_P:\t{partitions}~P~\t{of}~S}$, and it is straightforward to
work out that for any set $S$ of single-spin operators,

@@

  N!_{\abs{P}}\, O_P,

@@

While the expansions in \eqref{eq:multi_to_collective_full} and
\eqref{eq:multi_to_collective_full_simp} closely resemble each other,
in practice the latter involves less computational overhead due to the
lack of symmetrization in the construction of $\O_{\tilde P}$.  There
are, in principle, even more economical decompositions of a
multi-local operators into collective opreators, for example
\begin{align}
  N!_2\, X\otimes Y
  &\EQFS \col{X}\,\col{Y} - \col{XY}, \\
  N!_3\, X\otimes Y \otimes Z
  &\EQFS \col{X}\,\col{Y}\,\col{Z}
  - \col{X}\,\col{YZ} - \col{Y}\,\col{XZ} - \col{Z}\,\col{XY}
  + \col{ZXY} + \col{YXZ},
\end{align}
but these decompositions generally do not preserve the manifest
symmetry of a multi-local operator $O_S$ under arbitrary permutations
of $S$, and we do not know of any general prescription to arrive at
such decompositions aside from solving systems of equations relating
collective operators to multi-local operators.

On a final note, we consider the special case that all operators
$O\in S$ are identical, for which we provide the expansions
\begin{align}
  N!_2\, X\otimes X \EQFS {\col{X}\,}^2 - \col{X^2},
  &&
  N!_3\, X^{\otimes 3} \EQFS {\col{X}\,}^3
  - 3 \col{X} \odot \col{X^2} + 2\col{X^3},
\end{align}
\begin{align}
  N!_4\, X^{\otimes 4}
  \EQFS {\col{X}\,}^4 - 4 {\col{X}\,}^2 \odot \col{X^2}
  - 2 \col{X} \, {\col{X}\,}^2 \, \col{X}
  + 3 {\col{X^2}\,}^2 + 8 \col{X} \odot \col{X^3}
  - 6 \col{X^4}.
\end{align}
If furthermore $X^2=1$, as with e.g.~the Pauli operators of SU(2),
then
\begin{align}
  N!_2\, X\otimes X \EQFS {\col{X}\,}^2 - N,
  &&
  N!_3\, X^{\otimes 3} \EQFS {\col{X}\,}^3 - \p{3N-2} \col{X},
\end{align}
\begin{align}
  N!_4\, X^{\otimes 4} \EQFS {\col{X}\,}^4
  - 2\p{3N-4} {\col{X}\,}^2 + 3 N \p{N-2}.
\end{align}

>>>>

\newcommand{\FS}{\text{FS}}
\newcommand{\EQFS}{=_{\text{FS}}}
\newcommand{\col}{\underline}

\DeclareMathOperator{\diag}{diag}
\newcommand{\ul}{\underline}

@@

We now return to the task of computing the second-order effective
Hamiltonian $H_1^{(2)}$.  Any coefficient vector $\v v_X$ can be
expanded into its projections $\v v_X^\Delta$ onto the eigenspace of
vectors satisfying \eqref{eq:cond_1} with eigenvalue $\Delta$,
i.e.~$\v v_X = \sum_\Delta \v v_X^\Delta$, which also allows us to
expand
\begin{align}
  \V_1 = \sum_\Delta \V_1^\Delta,
  &&
  \V_1^\Delta \equiv \sum_X \v v_X^\Delta \c \v X,
\end{align}
where each operator $\V_1^\Delta$ generates a state with definite
interaction energy $\Delta$ above that of the fully symmetric manifold
$\M_0$.  We can therefore simplify
\begin{align}
  H_1^{(2)}
  = - \P_0 \V_1 \E \V_1 \P_0
  = - \sum_{\Delta\ne0} \f1{\Delta}
  \P_0 \V_1^\Delta \P_\Delta \V_1^\Delta \P_0
  = - \sum_{\Delta\ne0} \f1{\Delta} \P_0 \p{\V_1^\Delta}^2 \P_0,
\end{align}
where the product $\P_0 \p{\V_1^\Delta}^2 \P_0$ is worked out in
Appendix \ref{sec:PXYP}.  In total, we find
\begin{align}
  H_1^{(2)}
  \EQFS \f1{N\p{N-1}} \sum_{X,Y} \sum_{\Delta\ne0}
  \f{\v v_X^\Delta\c\v v_Y^\Delta}{\Delta}
  \p{\col{X}\,\col{Y} - N \col{XY}},
  \label{eq:H_1_2}
\end{align}
where $\EQFS$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$; $\v v_X^\Delta$ is the projection of
$\v v_X$ onto the $\Delta$-eigenspace of the matrix $\m h-\diag\v h$;
and for any single-body operator $Z$, we define the collective operator
\begin{align}
  \col{Z} \equiv \sum_p Z_p.
\end{align}

@@

Calculating the second-order effective Hamiltonian $H_1^{(2)}$ induced
on the fully symmetric manifold $\M_0$ by the perturbation $\V_1$ in
\eqref{eq:perturbations} requires us to simplify
\begin{align}
  \P_0 \p{\V_1^\Delta}^2 \P_0
  \EQFS \sum_{\substack{X,Y\\p,q}} v_{Xp} v_{Yq} X_p Y_q
  = \sum_{X,Y,p} v_{Xp} v_{Yp} X_p Y_p
  + \sum_{\substack{X,Y\\p\ne q}} v_{Xp} v_{Yq} X_p Y_q,
  \label{eq:PXYP_start}
\end{align}
where $\EQFS$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$; and all coefficient vectors $\v v_X$
satisfy \eqref{eq:cond_1} with eigenvalue $\Delta\ne0$, which implies
that all $\v v_X$ are mean-zero.  The first sum on the right of
\eqref{eq:PXYP_start} is then
\begin{align}
  \sum_{X,Y,p} v_{Xp} v_{Yp} X_p Y_p
  \EQFS \sum_{X,Y,p} v_{Xp} v_{Yp} X_0 Y_0
  \EQFS \f1N \sum_{X,Y} \v v_X \c\v v_Y \col{XY},
  &&
  \col{Z} \equiv \sum_p Z_p.
  \label{eq:PXYP_eq}
\end{align}
The second sum on the right of \eqref{eq:PXYP_start}, meanwhile, is
\begin{align}
  \sum_{\substack{X,Y\\p\ne q}} v_{Xp} v_{Yq} X_p Y_q
  \EQFS \sum_{\substack{X,Y\\p\ne q}} v_{Xp} v_{Yq} X_0 Y_1,
  \label{eq:PXYP_neq_start}
\end{align}
where we can use the fact that the vectors $\v v_X,\v v_Y$ are
mean-zero to simplify the sum over $p\ne q$ in
\eqref{eq:PXYP_neq_start} to
\begin{align}
  \sum_{p\ne q} v_{Xp} v_{Yq}
  = \sum_{p,q} v_{Xp} v_{Yq} - \sum_{p=q} v_{Xp} v_{Yq}
  = - \v v_X\c\v v_Y,
\end{align}
In order to simplify the product of operators in
\eqref{eq:PXYP_neq_start}, we expand
\begin{align}
  \col{X}\,\col{Y} = \sum_{p,q} X_p Y_q
  = \sum_{p=q} X_p Y_q + \sum_{p\ne q} X_p Y_q
  \EQFS \col{XY} + N\p{N-1} X_0 Y_1,
  \label{eq:PXYP_neq_ops}
\end{align}
which implies that
\begin{align}
  \sum_{\substack{X,Y\\p\ne q}} v_{Xp} v_{Yq} X_p Y_q
  \EQFS - \sum_{X\ne Y} \v v_X\c\v v_Y X_0 Y_1
  \EQFS - \f1{N\p{N-1}}\sum_{X,Y} \v v_X\c\v v_Y
  \p{\col{X}\,\col{Y} - \col{XY}}.
  \label{eq:PXYP_neq}
\end{align}
Altogether, we have that
\begin{align}
  \P_0 \p{\V_1^\Delta}^2 \P_0
  = - \f1{N\p{N-1}} \sum_{X,Y} \v v_X \c\v v_Y
  \p{\col{X}\,\col{Y} - N \col{XY}}.
\end{align}

@@

  \ut{\check{\m h}}

@@

  \ut{\v{\m w}} \equiv \sum_{d\ne0} w_{d,0} \ket{d},

@@

  \ut{\check{\m h}} \c \ut{\v{\m w}}
  = \p{\f12\Delta + h_0} \ut{\v{\m w}}.

@@

$\ut{\v{\m w}}$ as

@@

  \ut{\v{\m w}} = \sum_{d\in\VV_N} w_{d,0} \,

@@

Finally, we project $\ut{\check{\m h}}$ onto the reflection-symmetric

@@

  \P_+ \ut{\check{\m h}} \P_+

@@

  \P_+ \ut{\check{\m h}} \P_+ \c \ut{\v{\m w}}
  = \p{\f12\Delta + h_0} \ut{\v{\m w}}.

@@

$\P_+ \ut{\check{\m h}} \P_+$ only matter when $N$ is even, in which

@@

$\P_+ \ut{\check{\m h}} \P_+$ by a factor of $1/\sqrt{2}$.

@@

where we underline a single-spin operator to denote its collective
version, i.e.~$\col{Q}\equiv\sum_j Q^{(j)}$, where $Q^{(j)}$ is an
operator that acts with the single-spin operator $Q$ on spin $j$ and
trivially on all other spins.  By construction, the set of collective
operators $\set{\O_P:\t{partitions}~P~\t{of}~S}$ is one-to-one to the
set of multi-local operators $\set{O_P:\t{partitions}~P~\t{of}~S}$,
and it is straightforward to work out that for any set $S$ of
single-spin operators,

@@

  N!_{\abs{P}} O_P,

@@

While the expansions in \eqref{eq:multi_to_collective_full} and
\eqref{eq:multi_to_collective_full_simp} closely resemble each other,
in practice the latter involves less computational overhead due to the
lack of symmetrization in the construction of $\O_{\tilde P}$.  There
are, in principle, even more economical decompositions of a
multi-local operators into collective opreators, for example
\begin{align}
  N!_2 X\otimes Y
  &\EQFS \col{X}\,\col{Y} - \col{XY}, \\
  N!_3 X\otimes Y \otimes Z
  &\EQFS \col{X}\,\col{Y}\,\col{Z}
  - \col{X}\,\col{YZ} - \col{Y}\,\col{XZ} - \col{Z}\,\col{XY}
  + \col{ZXY} + \col{YXZ},
\end{align}
but these decompositions generally do not preserve the manifest
symmetry of a multi-local operator $O_S$ under arbitrary permutations
of $S$, and we do not know of any general prescription to arrive at
such decompositions aside from solving systems of equations relating
collective operators to multi-local operators.

On a final note, we consider the special case that all operators
$O\in S$ are identical, for which we provide the expansions
\begin{align}
  N!_2 X\otimes X \EQFS \col{X}^2 - \col{X^2},
  &&
  N!_3 X^{\otimes 3} \EQFS \col{X}^3
  - 3 \col{X} \odot \col{X^2} + 2\col{X^3},
\end{align}
\begin{align}
  N!_4 X^{\otimes 4}
  \EQFS \col{X}^4 - 4 \col{X}^2 \odot \col{X^2}
  - 2 \col{X} \, \col{X}^2 \, \col{X}
  + 3 \p{\col{X^2}}^2 + 8 \col{X} \odot \col{X^3}
  - 6 \col{X^4}.
\end{align}
If furthermore $X^2=1$, as with e.g.~the Pauli operators of SU(2),
then
\begin{align}
  N!_2 X\otimes X \EQFS \col{X}^2 - N,
  &&
  N!_3 X^{\otimes 3} \EQFS \col{X}^3 - \p{3N-2} \col{X},
\end{align}
\begin{align}
  N!_4 X^{\otimes 4} \EQFS \col{X}^4
  - 2\p{3N-4} \col{X}^2 + 3 N \p{N-2}.
\end{align}
